### 2022-01-31 papers/2022-rewriting-popl/response.txt

> lines 220-225, most of the details mentioned in this paragraph remain mysterious.
> Why would "extra bottom-up passes" be needed?
> (Doesn't rewriting produce a normal form?)
> What is a "permitted" identifier?
> What is `seq`? etc.

We have apparently not made it clear that our rewriter performs only a *single* NbE-ordered pass over the term.
Much like NbE is capable of fully reducing a term to normal form in a single pass, we've found that nearly all of our rewriting and compilation can be done in just a single pass over the AST.
For the sake of convenience, we allow limited localized exceptions to this single-pass rule, primarily useful for cases where the RHS of one rewrite rule is expressed in terms of identifiers that allow immediate additional rewriting.
Multiple passes are supported by wrapping the rewriter in a fueled loop at the Gallina level, but neither Fiat Cryptography nor our microbenchmarks required such looping.

Our plugin emits on-the-fly an inductive type of identifiers which can be used in the AST.
Any identifiers not in this inductive type cannot appear in goals to be reified.
By default, this list is composed of the identifiers found in any rewrite rule mentioned.
The `with extra idents` construct allows extending this list so that more goals can be reified.
The `seq` identifier is `Coq.Lists.List.seq`, which shows up in the definition of `prefixSums` but not in any rewrite rule.
It might seem like such identifiers are useless, because they will not be mentioned in any rewrite rule, but the `with delta` generates rules even for identifiers which show up only in the `with extra idents` list, and in fact in this example the rule that says to reduce `seq` to normal form only when its applied to fully concrete (ground / closed) arguments is adequate.

> line 290, what is "the former"?

Soundness.
"The latter" would be "syntactic-well-formedness", both from the preceding sentence "The statements of rewrite rules are reified and soundness and syntactic-well-formedness lemmas are proven about each of them."

> line 292, "Automating this step allows rewrite rules to be proven in terms of their shallow embedding".
> I don't know what this means.

In engines like R_tac, users must provide proofs about the reified, syntactic form of rewrite rules.
This is a significant burden for users unfamiliar with the ins and outs of R_tac.
In our engine, users never need to see the underlying reflective engine, nor perform any reasoning about it, thanks to automation.

More concretely, a user merely needs to prove a lemma like `forall n, n + 0 = n`, not `forall n, Expr.Interp (expr.App (expr.App (expr.Ident expr.Z_add) n) (expr.Ident (expr.Literal 0))) = Expr.Interp n` nor `forall n, Expr.Wf n -> Expr.Wf (expr.App (expr.App (expr.Ident expr.Z_add) n) (expr.Ident (expr.Literal 0)))`.

> line 312, "limited support".
> Limited in what ways?

We support only prenex type-polymorphism in constants, and furthermore the types must always be concretely instantiated whenever the constant actually appears in an expression.
For example, we support `List.append`, which is polymorphic over the type of the list, but we only support it on concretely typed lists.
That is, we can use `List.append` to concatenate two lists of `nat`s, but we can't talk about lists whose types are unknown or variable.

> line 325, "we instead build our rewriter in Coq’s logic".
> Not sure what this means.
> Is your rewrite a program expressed in Gallina?
> Is it something else?

Yes, our rewriter is a Gallina program.
We are contrasting this with Aehlig et al.'s rewriter, which is an OCaml program.

> line 330, you adopt Maranget's approach, because you fear that a simpler approach would involve "duplicate work".
> Have you compared the two approaches in practice?

We're about 8x faster than the simpler approach.
(We didn't include any numbers or plots because we've only done a very rough comparison, and didn't consider it worth it to do extensive performance-testing here.)
This 8x can be broken down into a 4x factor and a 2x factor:
Using decision trees gives us a 4x speed-up over the naive approach, and furthermore allows us to partially evaluate the rewriting engine itself to specialize it to the particular rewriting rules.
(This partial evaluation is not easily possible without the decision tree; we needed to carefully craft our engine to be written in CPS subject to a few other constraints to allow such partial evaluation to be useful, and enabling partial evaluation on the naive approach would require an entirely different set of constraints on how we wrote the engine.
Furthermore, partial evaluation done naively on the simpler approach would result in a roughly quadratic code-size blowup with no gain: it would duplicate the list of rewrite rules for every identifier in the language without actually specializing the logic to the particular rewrite rules.)
Actually performing the early partial evaluation, so that the decision tree does not need to be evaluated when the rewriting engine is run, buys us another 2x speedup.

> line 335, "There are three steps to turn a set of rewrite rules into a functional program that takes in an expression and reduces according to the rules."
> Is the goal even well-defined?
> What if the set of rewrite rules is not confluent or not terminating?

Here is another place where we would have benefited from being clearer that the goal is a *single* NbE-ordered pass over the term, which is nearly sufficient for the vast majority of our use cases.
Handling of non-confluent rewrite rules is determined first by NbE-order term traversal, and subsequently by applying all rules with the same head identifier in order and taking the first one that matches.
Any looping must be fueled, and non-terminating rules would run repeatedly until fuel runs out.

> line 379, what is "Coq's logic"?
> What is "Coq's normal partial evaluation"?

"Coq's logic" is Gallina.
"Coq's normal partial evaluation" is `cbv` / `lazy` / `cbn` with a whitelist or blacklist.

> line 440, is this an amendment to the code in Figure 1?
> It would be clearer to have just one version of the code.

Yes.
Thanks for the suggestion.

> line 482, I don't understand the implied opposition between "closed programs" and "thousands of lines of low-level code".
> One aspect is whether the program has free variables; another aspect is the size of the program.

In our experience, most routines scale super-linearly in the number of binders.
Hence a rewriting engine or reduction procedure that scales linearly in the size of the output term, in the absence of binders, might scale quadratically or even cubically in the number of binders in the output term.
This behavior will be invisible when the engine is only tested on closed programs and on open programs that produce only a couple lines of code.

> line 496, "it is difficult to integrate arrays soundly in a logic".
> This is a naïve question, but could Coq's primitive arrays be used here?

Coq's primitive arrays have no recursion scheme and, as far as we're aware, have not been used in any major development.
(Additionally, Coq had no primitive arrays when we completed work on this project.)
We would be interested in seeing a performance analysis of various ways of doing binder bookkeeping that included primitive arrays.

> line 504, "Here is the actual inductive definition".
> Could this definition come earlier?
> The definition of reify & reflect would be clearer if it did.

It could, though we were wary of overwhelming the reader with large amounts of code.

> line 517, "A good example of encoding adequacy".
> I believe I know what "adequacy" means in this context, but I don't think every reader will.

Noted, thanks.

> line 577, "we needed to convince Coq that the function terminates".
> Coq is convinced already, if it accepts your definition.
> The phrasing is somewhat odd.
>
> line 581, "the Coq kernel is ready to run our Rewrite procedure during checking."
> During checking of what?
> If you mean that the Coq type-checker performs reduction during type-checking, that is a well-known fact which perhaps does not need to be recalled here.

Indeed, the phrasing here is odd, and we'll reword it.

> line 584, "These strategies have".
> Not clear which strategies are meant here.

The reduction strategies built in to Coq's kernel which are referred to at the end of the previous sentence, "[...] it is important to note how the Coq kernel builds in reduction strategies."
In particular, we are referring to `lazy`, `vm_compute`, and `native_compute`.

> line 590, "“known-good” goals".
> I have no idea what this means.

We mean goals that line up with the sorts of goals on which Coq's built-in reduction strategies have been tuned.
Granted, this knowledge is probably only known to expert Coq users, and perhaps we should rephrase.

> line 616, "Then we modified reduction to inline let binders instead of preserving them, at which point the reduction job terminated with an out-of-memory error [...]".
> I don't really get the point.
> Even if the compiler had been able to terminate, the resulting "optimized" code would be practically unusable, as it would perform many redundant computations.
> Right?

That's right.
This is yet another argument that subterm sharing is crucial, which is what we're trying to argue in this section.

> the "NbE strategy that outputs UnderLets telescopes" in Section 4.2 is never fully defined, as far as I can tell.

The UnderLets type is defined on lines 639--641.
This type is a monad, and it is straightforward to sprinkle the monadic `bind` and `return` operators in the NbE code given from lines 546--567 in such a way that the code typechecks.
There is very little of note in the interplay of this monad with NbE, for all that it seems quite essential to the performance of our rewriting engine, and we felt that including this code in our initial presentation of NbE would overwhelm the reader with too many unfamiliar moving parts.

> line 650, why is `with_lets` needed?
> Is it used to implement the heuristics discussed on line 658?

No, it's unrelated to the heuristic on line 658.
Strictly speaking, `with_lets` is not needed.
However, including `with_lets` allows us to avoid monadic binds in a number of places, which simplifies some of the code.
Perhaps we should have dropped `with_lets` entirely from our presentation, since even our additional thirty pages of written content on the various design decisions, tradeoffs, and lessons does not delve into the tradeoffs of having `with_lets` vs avoiding it.
However, we felt that it was better to present exactly the code that we actually use, in this instance.

> I believe the authors could perhaps better explain why the rewriting rule at line 706 is preferable (somehow more tractable) than the rewriting rule at line 696.
> At first sight, they seem similar.
> I guess the point is that the variable `n` is unknown (i.e., *not* a compile-time constant) whereas the variable `u` is known (a compile-time constant), so the second rule falls within the scope of the technology described in Section 4.3, whereas the first rule does not.

Yes, this is correct.

> The `clip` trick seems to be a nice way of attaching statically computable information to variables.
> It may be worth discussing how general this trick is (can it be used to carry arbitrary information, not just an interval?) and whether there might be other ways of achieving the same result (could the information be stored in some sort of environment?).

Thanks!
Yes, it can carry arbitrary information.
However, we've neither needed nor evaluated it on any other sort of information, and expect that the reader can think up uses just as well as we can.
Rewrite rules don't have access to any sort of environment, though I suppose that section 5.1.6 can be seen as suggesting that we provide rewrite rules with a sort of environment of abstract information.

> lines 721 and 725, "Recall from section 2 [...]".
> This was never really explained.
> All of Section 4.5 is relatively obscure.

Explaining the mechanism by which the rewriter makes use of these markers is far beyond the scope of this paper, and including all such details of the rewriting engine would probably result in upwards of a hundred pages of content.
We were not sure how to make this section less obscure; these features implement relatively straightforward capabilities with highly technical code, and it's hard to say much more than "our framework supports this feature" without ending up exposing the reader to a great deal of detailed implementation minutiae.

> line 736, "when converting the interpretation of the reified term with the prereified term".
> I don't know what this means.

As per your description of our engine, the user gives a program in Gallina and asks for the optimized version.
This optmized version is equal to the original term.
Since our rewriting engine operates on deeply-embedded syntax, we must equate the user-given program in Gallina with the interpretation of its own reification.
This equality is proven by reflexivity, as the two terms are convertible, and we are referring to the user-given Gallina program as "the prereified term".

> line 755, I don't understand the meaning of "iterating on the Fiat Cryptography compiler".

Fiat Cryptography development is ongoing.
Porting the core of the compiler to this framework did not end development.
Some of the development since then has taken the form of tweaking the set of rewrite rules and rewriter passes, and we are referring to this development as "iterating on the Fiat Cryptography compiler".

> Section 5.1.2 seems relatively uninteresting.

Perhaps, though there's a sense in which section 5.1.2 might be more properly labeled "New Constructs in the Frontend and Intermediate Representations", and this easy ability to change the IRs seems to be something you argued is important above, when discussing lines 29-31.

> line 808, "designating Z.combine_at_bitwidth as an identifier that should be inlined rather than let-bound".
> I am not sure at which point it should be inlined away.
> Certainly not too soon, otherwise the rewrite rules mentioned at lines 805-806 would never fire.

Inlining occurs after all applicable rewrite rules have triggered at a location, and rewrite rules are allowed to let-bind subterms.
However, you may be misinterpreting what we mean by "inlined", since inlining would not prevent the rewrite rules mentioned at lines 805-806 from firing.
We mean that whenever the rewriter would emit a term such as `let x := Z.combine_at_bitwidth y in z`, it should instead emit `[Z.combine_at_bitwidth y/x]z`.

> line 834-847, I think I understand this paragraph (more or less), but I do not see how it explains the quartic behavior discussed in the previous paragraph.

The particular algorithm that we use contains quartically many variable assignments, but only quadratically many of them are distinct.
However, this constant folding combined with limited common subexpression elimination (realzing that you don't need to have a variable holding `x + 0` if you already have a variable `x`) can only be realized after bounds analysis.
The paragraph on lines 834-847 is a simplified example aimed at making concrete how the fusion can impact the number of operations used, rather than being precisely accurate about the source of quartic behavior.

> line 852, the forward pointer to section 5.2.4 is somewhat unfortunate; the text cannot be fully understood without looking ahead, it seems.
> We do not know what the parameters `n` and `m` represents.

Indeed, we should rearrange the text so that we're not referring to variables before we display the code.

> On line 853, it is not clear what "cubic" means (cubic in what parameter?).

Cubic in the same variable that the first half of the sentence expressed linear performance in terms of, `n · m`:
"We achieve linear performance in n · m when ignoring the final `cbv`, while `setoid_rewrite` and `rewrite_strat` are both cubic"

> On line 858, what is `make`?
> What is `map_dbl`? The whole paragraph (848-861) is extremely technical.
> It does not clearly explain why it is the fundamentally difficult to perform let-lifting in an efficient way.

We will move this paragraph after Figure 5, where `make` is defined.
Hopefully this will make the explanation clear, since all we are doing is arguing that a particular reduction strategy (performing all rewriting and reduction with the exception of let-lifing, followed by lifting the let binders, and repeating this sequence of two steps) has a particular asymptotic performance.

> It also does not explain its opening sentence, "Another essential pass to fuse with rewriting and partial evaluation is let-lifting".
> Why is it essential?
> Is it more efficient if fused than if performed separately?
> Why is that?

Setting aside the missing clarity of the proof of asymptotic behavior, which is hampered by the fact that this paragraph comes before Figure 5, we clearly state that performing let-lifting via rewrite rules results in $\mathcal O(n\cdot m^2)$ let-liftings, while we are able to achieve performance of $\mathcal O(n\cdot m)$ by fusing the passes.
The analysis is identical if let-lifting is performed in any other way, so long as it alternates with reduction rather than being fused with it.

> In Figure 2, is `m` in fact a known constant?
> Otherwise I am not sure what the ellipsis means.

For any particular data-point, both `m` and `n` are known constants.
However, we do evaluate the code for multiple values of `m`.

> line 958 (and elsewhere), "continuation-passing style [...] allows standard VM-based reduction to achieve good performance".
> Is it folklore knowledge why writing in CPS style helps?
> It might be worth briefly recalling why this is so.

This is alluded to on lines 619--636.
In the expression `map (λ x. y + x) (let z := e in [0; 1; 2; z; z + 1])`, the `map` will not reduce unless let-binders are inlined.
However, in the equivalent CPS'd expression `(λ K. map_cps (λ x K. K (y + x)) (λ K. let z := e in K [0; 1; 2; z; z + 1]) K) (λ x. x)` reduces to `let z := e in [y; y + 1; y + 2; y + z; y + (z + 1)]` as desired.
I'm not sure if there's any folklore citation to be had, and I'm also not sure how to express the general principle here.
